The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-27 article_final/main.tex
refactored abstract and introduction a little bit

\begin{abstract}
	In this article, we explore the use of various matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the duals to the Ky Fan $k$-norms to introduce a family of Muon-like algorithms we name \emph{Fanions}, which happen to be similar to Dion. Then working with the duals of convex combinations of the Ky Fan $k$-norms and the Frobenius norm or the $l_\infty$ norm, we construct the families of \emph{F-Fanions} and \emph{S-Fanions} respectively. Their most prominent members are \emph{F-Muon} and \emph{S-Muon}. We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that F-Muon and S-Muon are always on par with Muon, while on synthetic linear least squares they are even better than vanilla Muon optimizer.
\end{abstract}

@@

That is what can be said from a practical point of view. From a theoretical perspective, a key innovation of Muon was its principled derivation of the update rule, which emerged as the solution to an optimization problem constrained by the RMS-to-RMS norm (scaled version of the spectral norm) \citep{bernstein2025deriving}.

@@

\textit{In deriving Muon's update step, why should one constrain by the spectral or any other operator norm? How would alternative norms affect performance and computational cost?}

In this article, we tackle this question by actually showing that there are many viable non-operator norms. We leverage the family of norms dual to Ky Fan $k$-norms to derive a new family of \textbf{Fanions}, algorithms with low-rank updates. This approach theoretically explains the backbone of Dion's update \citep{ahn2025dion} and generalizes the memory-motivated application of the nuclear norm to Sharpness-Aware Minimization \citep{pethick2025sam}. As it was done with Muon, we come up with an effective procedure for computing Fanions' updates. Lanczos algorithm, which is described and compared with its competitors in \sectionref{sec:matrix-side}, is the most operation-effective algorithm, which, however, for now lacks an effective GPU- and PyTorch-friendly implementation.

Working with duals to conic combinations of dual norms, we construct the families of \textbf{F-Fanions} and \textbf{S-Fanions}, which are hybrids of Muon and NormalizedSGD and SignSGD, respectively.

Then we compare the performances of the algorithm families on various model and real-world problems:
\begin{itemize}
	\item Synthetic least squares experiment \sectionref{subsec:lls_exps}
	\item CIFAR-10 airbench \citep{cifar2023airbench}
	\item Pre-training NanoGPT on FineWeb dataset \citep{modded_nanogpt_2024}
	\item Fine-tuning NanoGPT on Tiny Stories \citep{eldan2023tinystoriessmalllanguagemodels}
\end{itemize}

Our experiments reveal important insights into the role of matrix norms in optimization. First, we show on the example of Neon, the one-rank Fanion, that not every LMO-based algorithm is effective, despite the same asymptotics in the general bounds of \citet{kovalev2025understanding} and \citet{riabinin2025gluon}. This suggests that existing theoretical guarantees should be reworked to explain empirical performance.

Most notably, our experiments on real-world tasks demonstrate that the choice of underlying matrix norm is remarkably flexible. On CIFAR-10 airbench, properly-tuned F-Muon and S-Muon achieve $94.02 \pm 0.13\%$ and $94.03 \pm 0.13\%$ accuracy, matching Muon's $94.01 \pm 0.13\%$ performance. On NanoGPT pre-training, F-Muon achieves 3.281 cross-entropy loss, while fully-tuned Muon achieves 3.279. Finally, S-Muon matches Muon on fine-tuning of NanoGPT on Tiny Stories, while F-Muon is far more resistant to the learning rate choice than Muon. These results show that Muon-like algorithms can maintain competitive performance even when the underlying norm constraint is significantly modified, answering affirmatively the central question posed above. Moreover, the tools from \sectionref{sec:conic_combo} give the researchers an unheard-of flexibility in designing algorithms that do not have to be modifications of Muon.

@@

\begin{table}[ht]
\floatconts
  {tbl:newton_schulz_reference}
  {\caption{Newton–Schulz iteration on the same matrices (for reference).}}
  {%
    \begin{tabular}{|c|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Matrix size    & Time (s) & Matvecs & Iterations & $err_1$ \\
      \arrayrulecolor{black}\hline
      $500 \times 500$     & 0.041    & 27 000  & 27         & 4.8e-3  \\
      $5000 \times 5000$   & 26.4     & 290 000 & 29         & 6.5e-3  \\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

@@

\textbf{Improvements of Muon.} Regarding the first point, a large number of applications and improvements of Muon has been proposed in less than a year. \citet{liu2025muon} adapted the algorithm for training language models larger than NanoGPT. \citet{shah2025practical} organized efficient hyperparameter transfer by combining Muon with maximal update parametrization. To construct their COSMOS optimizer, \citet{chen2025cosmoshybridadaptive} applied computationally intensive updates of SOAP optimizer \citep{vyas2025soap} to a low-dimensional ``leading eigensubspace'' while using memory-efficient methods like Muon for the remaining parameters. \citet{amsel2025polar, grishina2025chebyshev} proposed more efficient alternatives to Newton-Schulz algorithm. \citet{si2025adamuon} introduced AdaMuon which combines element-wise adaptivity with orthogonal updates. We suppose that the described above or similar techniques can be applied to our optimizers as well. For example, F-Muon and S-Muon also benefit from faster alternatives to Newton-Schulz iterations, and Fanions may be a great substitute to Muon in COSMOS, because, as we have shown in \sectionref{sec:matrix-side}, Lanczos algorithm is much faster than Newton-Schulz iterations on very large matrices.

\textbf{Theory behind Muon.} Regarding the second point, there has been a prolonged gap in theory behind Muon, simplistic derivation of \citet{bernstein2025deriving} based on \citet{bernstein2024oldoptimizernewnorm} excluded. This gap, as it seems to us, is not even now completely closed. For example, \citet{kovalev2025understanding} has provided convergence guarantees of Muon and in various settings, from which, however, Muon's supremacy cannot be recovered. Indeed, although the obtained bounds depend on the norm choice, the asymptotics of the convergence remain the same as for NSGD and other optimizers, $K = \cO(\epsilon^{-4})$ in an $L$-smooth stochastic case.

@@

\textbf{The LMO and Error Feedback.} We already mentioned that rank-$k$ unsharded Dion without error feedback and scaling of the update is Fanion-$k$. Since error feedback for Dion is very important as discovered by the ablation study in \citet{ahn2025dion}, F-Fanions and S-Fanions will benefit from it as well. In the context of federated learning, error feedback is effective even for compressed Muon \citep{gruntkowska2025efmuon}. The advantage of Fanions and S-Fanions is that less bits are required for transmission: $\sum_{i=1}^k \vu_i \vv_i^\top$ is easily transmitted as $\{\vu_1, \vv_1, \dots, \vu_k, \vv_k\}$ ($(m+n)\times k$ floats), while the sign part is as usual coded in $m \times n$ bits. Thus, the compression is ``built-in''. Moreover, there is an intriguing possibility to construct differentially private Fanions and S-Fanions with the specific more optimal non-Gaussian noise, as was done with DP-signSGD \citep{jang2024dplogsign}. This we leave for future research.

\textbf{The nuclear norm in the LMO.} As we found out only when transforming our results into an article, the nuclear norm has already been explored in the context of the linear minimization oracle. \citet{pethick2025sam} applied it to create $\nu$SAM, a new sharpness-aware minimization technique. It would be interesting to substitute $\normn{\cdot}$ with $\normkfk{\cdot}^\dagger$ in their approach. Since the SAM-neighborhood becomes more diverse, $k > 1$ might add to the accuracy boost, while preserving a small memory footprint and a small time overhead, if Dion-style power iterations are used.

\textbf{Orthogonal research.} Fanions, F-Fanions, and S-Fanions benefit from the general theoretical description of \citet{riabinin2025gluon}, and better learning rates can be predicted by calculating the trajectory smoothness. They could be transformed to Drop-Fanions by updating only some layers as in \citet{gruntkowska2025dropmuon}. They could be viewed as approximations of the Non-Euclidean Broximal Point Method for the corresponding norms \citep{gruntkowska2025noneuclideanbroximal}. One can clip them to produce ClippedScion-like algorithms \citep{pethick2025generalizedgradientnormclipping}. They could be made more memory-effective by the zero-order techniques that worked for Muon \citep{petrov2025zeromuon}. Finally, the results from \citet{shulgin2025inexactmuon} can be used to explain the robustness of Muon to the norm change that occurred in our experiments and to theoretically derive faster yet working approximate schemes to calculate the LMO; power iterations with a very limited number of iterations is a good algorithm to consider for the analysis.

>>>>

\begin{abstract}
	In this article, we explore the use of various matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm underlying the Muon update, we leverage duals of the Ky Fan $k$-norms to introduce a family of Muon-like algorithms we name \emph{Fanions}, which are closely related to Dion. By working with duals of convex combinations of the Ky Fan $k$-norms with either the Frobenius norm or the $l_\infty$ norm, we construct the families of \emph{F-Fanions} and \emph{S-Fanions}, respectively. Their most prominent members are \emph{F-Muon} and \emph{S-Muon}. We complement our theoretical analysis with an extensive empirical study of these algorithms across a wide range of tasks and settings, demonstrating that F-Muon and S-Muon consistently match Muon's performance, while outperforming vanilla Muon on synthetic linear least squares problems.
\end{abstract}

@@

From a theoretical perspective, a key innovation of Muon was its principled derivation of the update rule, which emerged as the solution to an optimization problem constrained by the RMS-to-RMS norm (a scaled version of the spectral norm) \citep{bernstein2025deriving}.

@@

\emph{In deriving Muon's update step, why should one constrain by the spectral or any other operator norm? How would alternative norms affect performance and computational cost?}

In this article, we tackle this question by showing that there are many viable non-operator norms. We leverage the family of norms dual to Ky Fan $k$-norms to derive a new family of \emph{Fanions}, algorithms with low-rank updates. This approach theoretically explains the backbone of Dion's update \citep{ahn2025dion} and generalizes the memory-motivated application of the nuclear norm to Sharpness-Aware Minimization \citep{pethick2025sam}. As was done with Muon, we develop an effective procedure for computing Fanions' updates. The Lanczos algorithm, which is described and compared with its competitors in \sectionref{sec:matrix-side}, is the most operation-efficient algorithm, though it currently lacks an effective GPU- and PyTorch-friendly implementation.

Working with dual norms and various convex combinations of norms, we construct the families of \emph{F-Fanions} and \emph{S-Fanions}, which are hybrids of Muon with NormalizedSGD and SignSGD, respectively.

We then compare the performance of these algorithm families on various model and real-world problems:
\begin{itemize}
	\item Synthetic least squares experiment \sectionref{subsec:lls_exps}
	\item CIFAR-10 airbench \citep{cifar2023airbench}
	\item Pre-training NanoGPT on FineWeb dataset \citep{modded_nanogpt_2024}
	\item Fine-tuning NanoGPT on Tiny Stories \citep{eldan2023tinystoriessmalllanguagemodels}
\end{itemize}

Our experiments reveal important insights into the role of matrix norms in optimization. First, using the example of Neon (the rank-one Fanion), we show that not every LMO-based algorithm is effective, despite sharing the same asymptotics in the general bounds of \citet{kovalev2025understanding} and \citet{riabinin2025gluon}. This suggests that existing theoretical guarantees should be refined to better explain empirical performance.

Most notably, our experiments on real-world tasks demonstrate that the choice of underlying matrix norm is remarkably flexible. On CIFAR-10 airbench, properly-tuned F-Muon and S-Muon achieve $94.02 \pm 0.13\%$ and $94.03 \pm 0.13\%$ accuracy, matching Muon's $94.01 \pm 0.13\%$ performance. On NanoGPT pre-training, F-Muon achieves 3.281 cross-entropy loss, while fully-tuned Muon achieves 3.279. Finally, S-Muon matches Muon on fine-tuning NanoGPT on Tiny Stories, while F-Muon is far more resistant to learning rate choice than Muon. These results show that Muon-like algorithms can maintain competitive performance even when the underlying norm constraint is significantly modified, providing an affirmative answer to the central question posed above. Moreover, the tools from \sectionref{sec:conic_combo} give researchers considerable flexibility in designing algorithms that need not be direct modifications of Muon.

@@

\begin{table}[ht]
\floatconts
  {tbl:newton_schulz_reference}
  {\caption{Newton-Schulz iterations on the same matrices (for reference).}}
  {%
    \begin{tabular}{|c|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Matrix size    & Time (s) & Matvecs & Iterations & $err_1$ \\
      \arrayrulecolor{black}\hline
      $500 \times 500$     & 0.041    & 27 000  & 27         & 4.8e-3  \\
      $5000 \times 5000$   & 26.4     & 290 000 & 29         & 6.5e-3  \\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

@@

\paragraph{Improvements of Muon} Regarding the first point, a large number of applications and improvements of Muon has been proposed in less than a year. \citet{liu2025muon} adapted the algorithm for training language models larger than NanoGPT. \citet{shah2025practical} organized efficient hyperparameter transfer by combining Muon with maximal update parametrization. To construct their COSMOS optimizer, \citet{chen2025cosmoshybridadaptive} applied computationally intensive updates of SOAP optimizer \citep{vyas2025soap} to a low-dimensional ``leading eigensubspace'' while using memory-efficient methods like Muon for the remaining parameters. \citet{amsel2025polar, grishina2025chebyshev} proposed more efficient alternatives to Newton-Schulz algorithm. \citet{si2025adamuon} introduced AdaMuon which combines element-wise adaptivity with orthogonal updates. We suppose that the described above or similar techniques can be applied to our optimizers as well. For example, F-Muon and S-Muon also benefit from faster alternatives to Newton-Schulz iterations, and Fanions may be a great substitute to Muon in COSMOS, because, as we have shown in \sectionref{sec:matrix-side}, Lanczos algorithm is much faster than Newton-Schulz iterations on very large matrices.

\paragraph{Theory behind Muon} Regarding the second point, there has been a prolonged gap in theory behind Muon, simplistic derivation of \citet{bernstein2025deriving} based on \citet{bernstein2024oldoptimizernewnorm} excluded. This gap, as it seems to us, is not even now completely closed. For example, \citet{kovalev2025understanding} has provided convergence guarantees of Muon and in various settings, from which, however, Muon's supremacy cannot be recovered. Indeed, although the obtained bounds depend on the norm choice, the asymptotics of the convergence remain the same as for NSGD and other optimizers, $K = \cO(\epsilon^{-4})$ in an $L$-smooth stochastic case.

@@

\paragraph{The LMO and Error Feedback} We already mentioned that rank-$k$ unsharded Dion without error feedback and scaling of the update is Fanion-$k$. Since error feedback for Dion is very important as discovered by the ablation study in \citet{ahn2025dion}, F-Fanions and S-Fanions will benefit from it as well. In the context of federated learning, error feedback is effective even for compressed Muon \citep{gruntkowska2025efmuon}. The advantage of Fanions and S-Fanions is that less bits are required for transmission: $\sum_{i=1}^k \vu_i \vv_i^\top$ is easily transmitted as $\{\vu_1, \vv_1, \dots, \vu_k, \vv_k\}$ ($(m+n)\times k$ floats), while the sign part is as usual coded in $m \times n$ bits. Thus, the compression is ``built-in''. Moreover, there is an intriguing possibility to construct differentially private Fanions and S-Fanions with the specific more optimal non-Gaussian noise, as was done with DP-signSGD \citep{jang2024dplogsign}. This we leave for future research.

\paragraph{The nuclear norm in the LMO} As we found out only when transforming our results into an article, the nuclear norm has already been explored in the context of the linear minimization oracle. \citet{pethick2025sam} applied it to create $\nu$SAM, a new sharpness-aware minimization technique. It would be interesting to substitute $\normn{\cdot}$ with $\normkfk{\cdot}^\dagger$ in their approach. Since the SAM-neighborhood becomes more diverse, $k > 1$ might add to the accuracy boost, while preserving a small memory footprint and a small time overhead, if Dion-style power iterations are used.

\paragraph{Orthogonal research} Fanions, F-Fanions, and S-Fanions benefit from the general theoretical description of \citet{riabinin2025gluon}, and better learning rates can be predicted by calculating the trajectory smoothness. They could be transformed to Drop-Fanions by updating only some layers as in \citet{gruntkowska2025dropmuon}. They could be viewed as approximations of the Non-Euclidean Broximal Point Method for the corresponding norms \citep{gruntkowska2025noneuclideanbroximal}. One can clip them to produce ClippedScion-like algorithms \citep{pethick2025generalizedgradientnormclipping}. They could be made more memory-effective by the zero-order techniques that worked for Muon \citep{petrov2025zeromuon}. Finally, the results from \citet{shulgin2025inexactmuon} can be used to explain the robustness of Muon to the norm change that occurred in our experiments and to theoretically derive faster yet working approximate schemes to calculate the LMO; power iterations with a very limited number of iterations is a good algorithm to consider for the analysis.
